---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r read_data}
library(readr)
activity <- read_csv("activity.csv", col_types = cols())
```


## What is mean total number of steps taken per day?

#### Number of steps taken each day

```{r mean_step}
suppressMessages(library(dplyr))

steps_per_day <- activity %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps, na.rm = TRUE))

steps_per_day
```

#### Histogram of daily steps taken

```{r step_hist,fig.width=4,fig.height=3}
library(ggplot2)

qplot(x = steps, data = steps_per_day, binwidth = 1000)
```

#### Basic statistics of daily steps taken

```{r basic_step_stat}
median_step <- steps_per_day$steps %>% median() 
mean_step   <- steps_per_day$steps %>% mean()
```

- Median: `r median_step`
- Mean:   `r mean_step %>% round(digits = 2)`

## What is the average daily activity pattern?

#### Time series plot of steps average per interval

```{r interval_mean,fig.width=4,fig.height=3}
interval_mean <- activity %>% 
  group_by(interval) %>% 
  summarise(steps_mean = mean(steps, na.rm = TRUE))

qplot(x = interval, y = steps_mean, data = interval_mean, geom = "line")
```

#### Interval with the highest average steps

```{r active_interval}
most_active <- interval_mean %>% 
  arrange(desc(steps_mean)) %>% 
  slice(1)
```

Interval `r most_active$interval` had the highest average steps taken with
`r most_active$steps_mean %>% round(digits = 2)` steps.

## Imputing missing values

```{r NAs}
na_row <- activity$steps %>% is.na() %>% sum()
```

Number of rows that contain missing values for steps (`NA`) is `r na_row`.

I will replace the missing values with the average value for that interval.

```{r impute}
activity_imputed <- activity %>% 
  left_join(interval_mean, by = "interval") %>% 
  mutate(steps = ifelse(is.na(steps), steps_mean, steps)) %>% 
  select(-steps_mean)
```

#### Measuring the effect of imputing

```{r step_hist_imputed,fig.width=4,fig.height=3}
steps_per_day_imputed <- activity_imputed %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps, na.rm = TRUE))

qplot(x = steps, data = steps_per_day_imputed, binwidth = 1000)
```

```{r basic_step_stat_imputed,comment=""}
median_step_imputed <- steps_per_day_imputed$steps %>% median()
mean_step_imputed   <- steps_per_day_imputed$steps %>% mean()

impute_change <- tribble(
  ~stat,     ~pre_impute_val,  ~post_impute_val,
  "mean",    mean_step,        mean_step_imputed,
  "median",  median_step,      median_step_imputed
) %>% 
  mutate(diff = post_impute_val - pre_impute_val) %>% 
  select(`Statistic` = stat, `Pre Impute Value` = pre_impute_val,
         `Post Impute Value` = post_impute_val, `Difference` = diff)

impute_change
```

After imputing, the mean increased considerably. Also the histogram became
more normalized. Missing values were decreasing the mean and median because some 
days had 0 steps in result of them.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday}
day_type_mean <- activity_imputed %>% 
  mutate(day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), 
                           "weekend",
                           "weekday")) %>% 
  mutate(day_type = as.factor(day_type)) %>% 
  group_by(interval, day_type) %>% 
  summarise(steps = mean(steps))

qplot(x = interval, y = steps, data = day_type_mean, geom = "line") +
  facet_grid(day_type ~ .)
```

Activity seems to start earlier in the week days.